---
title: "HW4 Tidy data and joins"
output: 
  github_document
---
## Pre-Work

#### Load packages

Load the necessary package (assuming you have installed them)
```{r message=FALSE}
library(dplyr)
library(gapminder)
library(ggplot2)
library(tidyr)
library(lattice)
library(countrycode)
```


## Problem 1: General data reshaping and relationship to aggregation.

#### - Activity #2

Make a tibble with one row per year and columns for life expectancy for two or more countries.
```{r  results='asis'}
tbl1.2 <- 
  gapminder %>%
  filter(country %in% c("United States", "China", "Canada")) %>%
  select(country, year, lifeExp)

tbl1.2 %>%
  head(15) %>%
  knitr::kable(digits = 2, align ="r", padding = 10)
#Due to the long table, we only display head 15 rows. 
```

We can easily make ggplot using the above table format, but that is not what the question is asking.
```{r}
tbl1.2 %>%
  ggplot(aes(x = year, y = lifeExp, color = country)) +
  geom_point() +
  ggtitle("Scatterplot of lifeExp vs year for country Canada, China, and United States")
```

The above table is a little bit hard to see, to let's spread it. 

```{r results='asis'}
tbl1.3 <- 
  tbl1.2 %>%
  spread(key = country, value = lifeExp)
tbl1.3 %>%
  knitr::kable(digits = 2, align ="r", padding = 10 )
```

Take advantage of this new data shape to scatterplot life expectancy for one country against that of another
```{r, fig.width=5.5, fig.height=5.5}
tbl1.3 %>%
  select(-year) %>%
  splom()
```


#### - Activity #3 Compute median of life expectancy for all possible combinations of continent and year. 

```{r  results='asis'}
tbl1.4 <- 
  gapminder %>%
  group_by(continent, year) %>%
  summarise(median = median(lifeExp))

tbl1.4 %>%
  head(15) %>%
  knitr::kable(digits = 2, align ="r", padding = 10)
```

We can easily make ggplot using the above table format, but I guess that is not what the question is asking.
```{r}
tbl1.4 %>%
  ggplot(aes(x = year, y = median, group = continent, color = continent)) +
  geom_line() + geom_point() + 
  ggtitle("The median of lifeExp trend for five continents")
```

The above table is a little bit hard to see, to let's spread it. 

```{r results='asis'}
tbl1.5 <- 
  tbl1.4 %>%
  spread(key = continent, value = median)
tbl1.5 %>%
  knitr::kable(digits = 2, align ="r", padding = 10)
```

#### Is there a plot that is easier to make with the data in this shape versis the usual form? Try it!

Pairwise scatterplot is easier to do using this format of data form. 

```{r, fig.width=9, fig.height=8}
tbl1.5 %>%
  select(-year) %>%
  splom()
```


## Problem 2: Join, merge, look up

#### - Activity #1 Create a second data frame, complementary to Gapminder. Join this with (part of) Gapminder using a dplyr join function and make some observations about the process and result. Explore the different types of joins.

We generated another dataframe containing all non-Asian countries with country code and country name in French, and deleting the rows that contains NA (i.e. the countries whose values were not matched unambiguously, and where the warning comes from), and eventually, we have a 107 rows * 4 columns dataframe.    
#### Create a new dataset

```{r}
ctryDat <- 
  gapminder %>%
  filter(continent != "Asia") %>%
  distinct(country) %>%
  select(country = `country`) %>%
  mutate(ctryCode = countrycode(country, 'country.name', 'iso3c'), 
         ctryCodeNum = countrycode(country, 'country.name', 'iso3n'),
         ctryFR = countrycode(country, 'country.name', 'country.name.fr')) %>%
  na.omit()
dim(ctryDat)

ctryDat %>%
  head(15) %>%
  knitr::kable(digits = 2, align ="r", padding = 10)
```

#### Explore left_join

```{r}
tbl2.1 <- 
  left_join(ctryDat, gapminder, by = "country")
dim(tbl2.1) 
```

Even though my first thought would be the left join, and we only want 107 rows (as the ctryDat), the left_join still gives me 1284 rows, which is 12 times of the # of rows than I expected. The reason is that because gapminder has 12 different years for the same country, so, if we don't specify the year, it would join all the years of country. Let's try anohter example. 

```{r}
ctryDat2<- 
  ctryDat %>%
  mutate(year = 2007)
tbl2.2 <- 
  left_join(ctryDat2, gapminder, by = c("country", "year"))
dim(tbl2.2) 

tbl2.2 %>%
  head(15) %>%
  knitr::kable(digits = 2, align ="r", padding = 10)
```

It can be seen that now the joining works fine as we expected. 

```{r}
tbl2.3 <- 
  left_join(gapminder, ctryDat2, by = c("country", "year"))
dim(tbl2.3) 

tbl2.3 %>%
  tail(15) %>%
  knitr::kable(digits = 2, align ="r", padding = 10)
```

When doing the left join above, the dataframe of gapminder will be kept, and adding the additional columns according to the joining conditions.


#### Explore inner_join()

```{r}
tbl2.4 <- 
  inner_join(ctryDat2, gapminder, by = c("country", "year"))

which((
  inner_join(ctryDat2, gapminder, by = c("country", "year")) %>%
    select(country, continent, year, lifeExp, pop, gdpPercap, ctryCode,
           ctryCodeNum, ctryFR) ==
  inner_join(gapminder, ctryDat2, by = c("country", "year")) %>%
    select(country, continent, year, lifeExp, pop, gdpPercap, ctryCode,
           ctryCodeNum, ctryFR)) != 
    TRUE)

tbl2.4 %>%
  head(15) %>%
  knitr::kable(digits = 2, align ="r", padding = 10)
```

From above, we can see that the order of the two dataset in inner join does not affect the joining effects, just rearrange the order of variables, but not affect the values within each variable, and does not affect the number of rows in the result. 

#### Explore full_join
```{r}
tbl2.5 <- 
  full_join(ctryDat2, gapminder, by = c("country", "year"))

dim(tbl2.5)

tbl2.5 %>%
  head(15) %>%
  knitr::kable(digits = 2, align ="r", padding = 10)
```

Full join will join the two dataframe all together, without throwing away any mismatching rows. And no matter which data frame you put in first, the results in terms of the content are the same. 

#### Explore semi_join()
```{r}
tbl2.6 <- 
  semi_join(gapminder, ctryDat2, by = c("country", "year"))

tbl2.6 %>%
  head(15) %>%
  knitr::kable(digits = 2, align ="r", padding = 10)
```

```{r}
tbl2.7 <- 
  semi_join(ctryDat2, gapminder, by = c("country", "year"))
dim(tbl2.7)
tbl2.7 %>%
  head(15) %>%
  knitr::kable(digits = 2, align ="r", padding = 10)
```
semi_join() returned to a subset of the dataframe positioned first, also deleting the rows do not match values. 

#### Explore anti_join()
```{r}
tbl2.8 <- 
  anti_join(ctryDat2, gapminder, by = c("country", "year"))
dim(tbl2.8)
```


```{r}
tbl2.9 <- 
  anti_join(gapminder, ctryDat2, by = c("country", "year"))
dim(tbl2.9)
tbl2.9 %>%
  head(15) %>%
  knitr::kable(digits = 2, align ="r", padding = 10)
```

I found that anti_join() and semi_join() can be complement to each other. The result of `tbl2.7` and `tbl2.9` consists the whole gapminder dataframe; Similarly, the result of `tbl2.6` and `tbl2.8` consists the whole ctryDat2 dataframe. 

## Report your process

As I have some experience with ggplot and dplyr. So I firstly tried to tackle the problems without reviewing any materials. However, when I tried the following code, I got an error. The error stucks me a few minutes, so I went back to the course materials, and found out I misplaced `filter` and `select`. I did not pay much attention to the difference before, but now I have clearer idea and won't make the same mistakes next time.  

```{r  eval=FALSE}
gapminder %>%
  select(country %in% c("United States", "China", "Canada")) %>%
  itler(country, year, lifeExp)
```

Error in country %in% c("United States", "China", "Canada"): object 'country' not found

In addition, I found it is not easy to construct an appropriate dataframe. My `ctryDat2` is not a perfect one in demonstrating the _join() types, because it does not have the country beyound the range of `gapminder` dataset. Also, gapminder is a very large dataset, so the _join() results may be not so straightforward to check. 