---
title: "HW3 Gapminder Exploration Using Package(dplyr)"
output: 
  github_document
---
## Pre-Work

#### Load packages

Load the necessary package (assuming you have installed them)
```{r message=FALSE}
library(dplyr)
library(gapminder)
library(ggplot2)
library(tidyr)
```

## 1. Get the maximum and minimum of GDP per capita for all continents.

```{r results='asis'}
tbl1 <- 
  gapminder %>% 
  group_by(continent) %>% 
  summarise(minGDP = min(gdpPercap), medianGDP = median(gdpPercap), maxGDP = max(gdpPercap))
knitr::kable(tbl1)
```

```{r}
tbl1 %>%
  ggplot(aes(x= continent, y = medianGDP, color=continent)) + 
  geom_crossbar(aes(ymin  = minGDP, ymax = maxGDP), position = "dodge", width = 0.45) +
  xlab("Continent") +
  ylab("GDP per capita") +
  ggtitle("Range of GDP per capita separated by continent")
```

## 2. Look at the spread of GDP per capita within the continents.

Look at the IQR
```{r results='asis'}
tbl2 <-
  gapminder %>%
  group_by(continent) %>%
  summarise(`25%Q`=quantile(gdpPercap, probs=0.25),
            `50%Q`=quantile(gdpPercap, probs=0.5),
            `75%Q`=quantile(gdpPercap, probs=0.75))
knitr::kable(tbl2)
```

```{r}
gapminder %>%
  ggplot(aes(x= continent, y = gdpPercap, color=continent)) + 
  geom_boxplot() +
  xlab("Continent") +
  ylab("GDP per capita") +
  ggtitle("Boxplot of GDP per capita separated by continent")
```

##3. Compute a trimmed mean of life expectancy for different years. Or a weighted mean, weighting by population. Just try something other than the plain vanilla mean.

```{r results='asis'}
# Weighted mean
gapminder %>%
  group_by(year) %>%
  mutate(weighted_lifeE = weighted.mean(lifeExp, pop)) %>%
  select(year, weighted_lifeE) %>%
  distinct(year, weighted_lifeE) %>%
  knitr::kable()
```
```{r}
gapminder %>%
  group_by(year) %>%
  mutate(totPop = sum(as.numeric(pop)), 
         weighted_lifeE = weighted.mean(lifeExp, pop)) %>%
  ggplot(aes(x = year, y = weighted_lifeE, size = totPop)) +
  geom_point()+
  xlab("Year") +
  ylab("Weighted life expectancy with the weight of population") +
  ggtitle("Population-weighted life expectancy changes over time")
```

```{r results='asis'}
# 10% Trimmed Mean: Remove the top and bottom 10% (#14) of data
gapminder %>%
  group_by(year) %>%
  mutate(trimed_mean = mean(lifeExp, trim = .1)) %>%
  distinct(year, trimed_mean) %>%
  knitr::kable()
```


## 4.How is life expectancy changing over time on different continents?
```{r}
tbl4 <- 
  gapminder %>%
  select(continent, year, lifeExp)
knitr::kable(head(tbl4))
```
```{r}
tbl4 %>%
  ggplot(aes(x = year, y = lifeExp)) + 
  facet_wrap(~continent) + 
  geom_point(alpha = 0.35, color = "dark blue") + 
  geom_smooth(method = "loess") +
  xlab("Year") +
  ylab("Life expectancy") +
  ggtitle("Life expectancy changes over time for each continent")
```


## 5. Report the absolute and/or relative abundance of countries with low life expectancy over time by continent
```{r results='asis'}
#BenchMark: median of lifeExp
benchmark <- 60
tbl5 <- 
  gapminder %>%
  group_by(continent, year) %>% 
  mutate(NumOfCountries = n(), shareOfThisCountry = 1/NumOfCountries) %>%
  filter(lifeExp < benchmark) %>%
  summarise(n = n(), perCen = sum(shareOfThisCountry))

knitr::kable(head(tbl5))
```
```{r}
tbl5 %>%
  ggplot(aes(x= year, y = perCen, group= continent, color= continent)) +
  geom_line() +
  xlab("Year") +
  ylab("Percentage of countries whose life expectancy below benchmark(60)") +
  ggtitle("Each continent's below-benchmark country percentage over time")
```

## 6. Make up your own: The absolute increase population
The population increase is `pop` of the current year minus previous year, and group by continent and see whether the year increase differ by the continent.
```{r  results='asis'}
delta <- function(x){
  x.lag <- lag(x)
  x.delta <- x - x.lag
  return(x.delta)
}

tbl6 <- 
  gapminder%>%
  group_by(continent, year) %>%
  summarise(totPop = sum(as.numeric(pop))) %>%
  mutate(delta = delta(totPop)) %>%
  na.omit()
knitr::kable(head(tbl6))
```


```{r}
tbl6 %>%
  ggplot(aes(x = year, y = delta, color = continent, group= continent)) +
  geom_line() + 
  geom_point() +
  xlab("Year") +
  ylab("Increase in the population") +
  ggtitle("The increase in population over time separated by continent")
```



## 7. Find countries with interesting stories. 
I compare the GDP increase percentage between two countries. It can be seen China has an increasing GDP increase per capita, and Japan has a decreasing increase per capita. 
```{r  results='asis'}
deltaPC <- function(x){
  x.lag <- lag(x)
  x.delta <- (x - x.lag)/x.lag
  return(x.delta)
}

tbl7 <- 
  gapminder %>%
  filter(country %in% c("China", "Japan"))%>%
  group_by(country) %>%
  mutate(gdpIncPC = deltaPC(gdpPercap)) %>%
  na.omit()
knitr::kable(tbl7)
```

```{r}
tbl7 %>%
  ggplot(aes(x = year, y = gdpIncPC, color = country)) + 
  geom_point() + 
  geom_smooth(method = "lm")+
  xlab("Year") +
  ylab("Increase in the GDP per capita") +
  ggtitle("Increase in the GDP per capita over time for country China and Japan")
```

